---
title: "Beta Regression with rstanarm"
output: html_notebook
---

Can I squeeze my problem into the standard specification offered by `rstanarm`? They're supposed to have coded their models pretty efficiently. I'll be following <a href=https://cran.rstudio.com/web/packages/rstanarm/vignettes/betareg.html>this vignette</a>.

The main issue would be that seasonality will need to be handled with a design matrix. I don't think we're inverting exog matrix anywhere, so maybe that's fine.

```{r, message=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = "~/src/LondonMirror/Prepayments/")
setwd("~/src/LondonMirror/Prepayments/")
library(tidyverse)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

library(rstanarm)

sample_data = read_csv("data/samples.csv") %>% 
                rename(pool_number=X1, as_of_date=X2)
#Scale it
sample_data <- sample_data %>% mutate(burnout = burnout * 1e-7, 
                                      cato = cato * 0.1, 
                                      next_month_cpr = next_month_cpr * 1e-2,
                                      incentive = incentive * 5e-5, 
                                      lockin = lockin * 200, 
                                      sato = sato * 1e-4, 
                                      upfront_mip = upfront_mip * 0.01)
#filter out negative CPRs
sample_data <- sample_data %>% filter(next_month_cpr >= 0)
```

...actually, from reading the vignette, I think I can just specify seasonality as `1 | month`, so I'm setting eval=F on the next chunk.

```{r, eval=FALSE}
sample_data <- sample_data %>% mutate(month_1 = ifelse(seasonality==1,1,0),
                                      month_2 = ifelse(seasonality==2,1,0),
                                      month_3 = ifelse(seasonality==3,1,0),
                                      month_4 = ifelse(seasonality==4,1,0),
                                      month_5 = ifelse(seasonality==5,1,0),
                                      month_6 = ifelse(seasonality==6,1,0),
                                      month_7 = ifelse(seasonality==7,1,0),
                                      month_8 = ifelse(seasonality==8,1,0),
                                      month_9 = ifelse(seasonality==9,1,0),
                                      month_10 = ifelse(seasonality==10,1,0),
                                      month_11 = ifelse(seasonality==11,1,0),
                                      month_12 = ifelse(seasonality==12,1,0))
```

```{r betareg}

fit <- stan_betareg(next_month_cpr ~ cato + sato + hpa + lockin + burnout + 
                      incentive + upfront_mip + 1 | seasonality, 
                    data=sample_data[1:2500,], link="logit", chains=4, 
                    cores=4, iter=1000, prior_intercept=normal(0,1),
                    prior=normal(0,1), prior_phi=normal(0,1))
round(coef(fit), 2)
```

Can't have 0 and 1, but 0 is really common. Cribari-Neto <a href=https://cran.r-project.org/web/packages/betareg/vignettes/betareg.pdf>says</a>: 
<blockquote>Furthermore, if y also assumes the extremes 0 and 1, a useful
transformation in practice is $(y \dot (n − 1) + 0.5)/n$ where n is the sample size (Smithson and
Verkuilen 2006).</blockquote>

What does that transformation look like?
```{r}
N <- nrow(sample_data)
sample_data %>% ggplot(aes(x=next_month_cpr,
                           y=(next_month_cpr*(N-1) + 0.5)/N)) +
  geom_point() + ggtitle("Normalize next_month_cpr")
```
Huh. Don't have an intuitive sense for why that was one-to-one.
```{r}
min((sample_data$next_month_cpr*(N-1) + 0.5)/N)
max((sample_data$next_month_cpr*(N-1) + 0.5)/N)
```

Anyway, that was the recommended transformation.

```{r betareg2}
sample_data <- sample_data %>% 
  mutate(next_month_cpr2 = (next_month_cpr*(N-1) + 0.5)/N)

fit <- stan_betareg(next_month_cpr2 ~ cato + sato + hpa + lockin + burnout + 
                      incentive + upfront_mip + 1 | seasonality, 
                    data=sample_data, link="logit", chains=4, 
                    cores=4, iter=1000, prior_intercept=normal(0,1),
                    prior=normal(0,1), prior_phi=normal(0,1))
round(coef(fit), 2)
```
Unless I'm interpreting beta regression incorrectly, most of these coefficients are opposite to what they should be. You have a large intercept and then items subtracting from it.

What does phi_intercept mean? I thought phi must be > 0.


They seem to be stable, moreover; I've run it a few times.

Do I have seasonality specified correctly?

How does it compare to my normal model? Need to run normal model again with explicitly generated loglikelihood to compare.

The betas for the normal model were:
(Intercept)              cato              sato 
             0.15             -0.08             0.09 
              hpa            lockin           burnout 
            0.0126             -0.01              -0.37 
        incentive       upfront_mip (phi)_(Intercept) 
            -0.225             -0.06
            
So that had both `burnout` and `incentive` facing the wrong way.

`burnout` and `incentive` are likely collinear.

Let's try a Beta regression written by hand by myself.
