---
title: "20170310 - Popova Model with Full Histories"
output: html_notebook
---

I'm back to using CPR for endo. 

Betas are:
* incentive: WAC - prevailing mortgage rate, as before
* spline: incentive cubed
* burnout: WALA (can't do the more complicated burnout as I don't have all the historical data)
* seasonality: 1 if it's the summer-time, i.e. May-August.
* yield_curve: 10Y rate minus 2Y rate

I want to use the vintage (i.e. origination year) to adjust the betas a bit, see if I can't get a positive beta for incentive (at least in some years) without constraining it.

First I'll just do a shrunk intercept, then I'll try a full multi-level beta model.

```{r, message=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = "~/src/LondonMirror/Prepayments/")
setwd("~/src/LondonMirror/Prepayments/")
library(tidyverse)
library(lubridate)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

sample_data = read_csv("/data/prepayments/vnm_samples.csv") %>%
  rename(dt=major, cusip=minor)

sample_data <- sample_data %>% filter(endo >= 0 & incentive < 0.5)
N <- nrow(sample_data)
sample_data <- sample_data %>% mutate(endo = ((N-1) * (endo * 1e-2) + 0.5)/N,
                                      incentive = incentive * 10, 
                                      spline = spline * 1.3e3, 
                                      yc = yc * 0.5,
                                      burnout=burnout * 0.33e-2)
```

How does the data look on this one?
```{r}
sample_data %>% sample_n(size = 1000) %>%
  ggpairs(mapping=list(alpha=0.1), columns=c("incentive","spline","burnout", "yc","dt", "vintage","endo"),
                        upper = list(continuous = wrap("density", alpha = 0.5), combo = "box_no_facet"))

```

Burnout looks much more obvious when you see vintage ~ endo. There's a clear sweet spot.
On this dataset, incentive would probably get a positive beta if I right-censor it, too.

Anyway, stick to the plan. First pooled intercept by vintage.

```{r shrunk_intercept}
stan_code = "data {
    int N; #Number of records
    int K; #number of betas
    int V; #number of vintages
    int vintage[N]; #Pool vintage
    matrix[N,K] exogs;
    real endo[N];
}
parameters {
    row_vector[K] beta;
    real intercept;
    real vintage_intercept[V];
    real<lower=0.1> lambda; #dispersion
}
transformed parameters {
    vector[N] phi; #mu
    for(n in 1:N) {
      phi[n] = inv_logit(intercept + 
                        vintage_intercept[vintage[n]] +
                        beta * exogs[n]');
    }
}
model {
  intercept ~ normal(0, 0.1);
  to_vector(vintage_intercept) ~ normal(0, 0.1);
  beta[1] ~ normal(1,1); #incentive
  beta[2] ~ normal(0,5); #spline
  beta[3] ~ normal(-1,1); #burnout
  beta[4] ~ normal(1,1); #yc
  beta[5] ~ normal(1,5); #is_summer
  lambda ~ pareto(0.1, 1.5); #as per Gelman, 2013, ch.5
  endo ~ beta(lambda*phi, lambda*(1-phi));
}
generated quantities {
  real log_lik[N];
  real endo_hat[N];
  for(n in 1:N) {
    log_lik[n] = beta_lpdf(endo[n] | lambda*phi[n], lambda*(1-phi[n]));
    endo_hat[n] = beta_rng(lambda*phi[n], lambda*(1-phi[n]));
  }
}

"
less_data <- sample_data %>% sample_n(size=5000) %>% 
    mutate(vintage=year(vintage) - min(year(vintage)) + 1)
stopifnot(all(diff(sort(unique(less_data$vintage))) == 1)) #ensure we don't have any unused parameters in the fit.
fit <- stan(model_code=stan_code, model_name="Vintage_shrunk_intercept",
            data=list(N=nrow(less_data), K=5, V=length(unique(less_data$vintage)),
                      exogs=less_data %>% select(incentive, spline, 
                                                 burnout, yc, seasonality) %>%
                            as.matrix,
                      vintage=less_data$vintage, endo=less_data$endo),
            iter=2000, chains=4, pars=c("phi", "shrunk_intercept"), include=FALSE)
```

```{r}
print(fit, pars=c("beta", "intercept", "vintage_intercept", "lambda"), include=T)
```
It's not finding much to work with here. I'm going to re-run with censored incentive.


```{r shrunk_intercept_censor_incentive}
stan_code = "data {
    int N; #Number of records
    int K; #number of betas
    int V; #number of vintages
    int vintage[N]; #Pool vintage
    matrix[N,K] exogs;
    real endo[N];
}
parameters {
    row_vector[K] beta;
    real intercept;
    real vintage_intercept[V];
    real<lower=0.1> lambda; #dispersion
}
transformed parameters {
    vector[N] phi; #mu
    for(n in 1:N) {
      phi[n] = inv_logit(intercept + 
                        vintage_intercept[vintage[n]] +
                        beta * exogs[n]');
    }
}
model {
  intercept ~ normal(0, 0.1);
  to_vector(vintage_intercept) ~ normal(0, 0.1);
  beta[1] ~ normal(1,1); #incentive
  beta[2] ~ normal(0,5); #spline
  beta[3] ~ normal(-1,1); #burnout
  beta[4] ~ normal(1,1); #yc
  beta[5] ~ normal(1,5); #is_summer
  lambda ~ pareto(0.1, 1.5); #as per Gelman, 2013, ch.5
  endo ~ beta(lambda*phi, lambda*(1-phi));
}
generated quantities {
  real log_lik[N];
  real endo_hat[N];
  for(n in 1:N) {
    log_lik[n] = beta_lpdf(endo[n] | lambda*phi[n], lambda*(1-phi[n]));
    endo_hat[n] = beta_rng(lambda*phi[n], lambda*(1-phi[n]));
  }
}

"
N <- nrow(less_data)
censored_data <- less_data %>% 
  mutate(incentive=sapply(incentive, function(x){min(0.35,x)}))
fit <-  stan(model_code=stan_code, model_name="Vintage_shrunk_intercept",
            data=list(N=N, K=5, V=length(unique(less_data$vintage)),
                      exogs=censored_data %>% select(incentive, spline,
                                                     burnout, yc, seasonality) %>%
                            as.matrix,
                      vintage=censored_data$vintage, endo=censored_data$endo),
            iter=2000, chains=4, pars=c("phi", "shrunk_intercept"), include=FALSE)
```

```{r}
print(fit, pars=c("beta", "intercept", "vintage_intercept", "lambda"), include=T)
```

Didn't help. On to shrunk betas.
```{r shrunk_beta}
stan_code = "data {
    int N; #Number of records
    int K; #number of betas
    int V; #number of vintages
    int vintage[N]; #Pool vintage
    matrix[N,K] exogs;
    real endo[N];
}
parameters {
    row_vector[K] beta;
    matrix[V,K] vintage_beta;
    real intercept;
    real vintage_intercept[V];
    real<lower=0.1> lambda; #dispersion
}
transformed parameters {
    vector[N] phi; #mu
    for(n in 1:N) {
      phi[n] = inv_logit(intercept + 
                vintage_intercept[vintage[n]] +
                (beta + vintage_beta[vintage[n]]) * exogs[n]');
    }
}
model {
  intercept ~ normal(0, 0.1);
  to_vector(vintage_intercept) ~ normal(0, 0.1);
  beta[1] ~ normal(1,1); #incentive
  beta[2] ~ normal(0,5); #spline
  beta[3] ~ normal(-1,1); #burnout
  beta[4] ~ normal(1,1); #yc
  beta[5] ~ normal(1,5); #is_summer
  to_vector(vintage_beta) ~ normal(0,1);
  lambda ~ pareto(0.1, 1.5); #as per Gelman, 2013, ch.5
  endo ~ beta(lambda*phi, lambda*(1-phi));
}
generated quantities {
  real log_lik[N];
  real endo_hat[N];
  for(n in 1:N) {
    log_lik[n] = beta_lpdf(endo[n] | lambda*phi[n], lambda*(1-phi[n]));
    endo_hat[n] = beta_rng(lambda*phi[n], lambda*(1-phi[n]));
  }
}

"
N <- nrow(less_data)
fit2 <-  stan(model_code=stan_code, model_name="Vintage_shrunk_beta",
            data=list(N=N, K=5, V=length(unique(less_data$vintage)),
                      exogs=less_data %>% select(incentive, spline,
                                                     burnout, yc, seasonality) %>%
                            as.matrix,
                      vintage=less_data$vintage, endo=less_data$endo),
            iter=2000, chains=4, pars=c("phi"), include=FALSE)
```
Surprisingly fast.

```{r}
print(fit2, pars = c("beta","vintage_beta","intercept","vintage_intercept"))
```

Too much to look at. I'm mainly interested in what it did to incentive over time.

```{r}
vintage_beta <- extract(fit2,pars=c("vintage_beta"))[["vintage_beta"]]
vintage_beta <- apply(vintage_beta,c(2,3),median)
dimnames(vintage_beta) <- list(1985+seq.int(nrow(vintage_beta)),
                               c("incentive","spline","burnout","yc","season"))
vintage_beta <- as.data.frame(cbind(vintage=as.numeric(rownames(vintage_beta)),vintage_beta))

global_beta <- apply(extract(fit2, pars=c("beta"))[["beta"]],2,median)
global_beta <- data.frame(global_beta,beta=colnames(vintage_beta)[-1])

vintage_beta %>% gather(beta,value,-vintage) %>% 
  ggplot(aes(x=vintage,y=value, col=beta)) + geom_line() + 
    geom_hline(data=global_beta, mapping=aes(yintercept=global_beta,col=beta)) +
    ggtitle("Betas by Vintage (global beta as line)")
```

So here you go. Incentive is causing refinancing for younger loans. Otherwise it seems to have a negative effect, especially directly after the crisis.

Your base problem is that *incentive* and *home price appreciation* are highly correlated in the period for which we have the greater mass of data. Additionally, you have *burnout*, i.e. the fact that nobody refinances in the first 30 months or so; again, this is still more significant when the preponderance of data we're using to estimate is from the later period. 

It's not working to estimate the effects as if they are independent. Here, I suppose, is where the literature starts talking about *turnover* versus *incentive* models. 

So, how to incorporate these effects? We need to model interactions. In a traditional OLS paradigm, you do that by multiplying the (scaled) exogs. We can try that first. Warrants a new notebook.