---
title: "20170413 - Looking at LoanLevel data"
output: html_notebook
---

I have 100,000 samples of similar data to the pool-level model, but at loan level. Let's have a look at it.

```{r}
knitr::opts_chunk$set(include = FALSE)
library(knitr)
opts_knit$set(root.dir = "~/src/LondonMirror/Prepayments/")
setwd("~/src/LondonMirror/Prepayments/")
library(tidyverse)
library(lubridate)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

sample_data = read_csv("/data/prepayments/loans.csv") 
```

There are 40 columns, so a regular splom is going to be difficult to interpret. Also, removal_reason is messed up. It should have come (I think), from the forward month, not from the current month. As it sits, all removal reasons are blank.

In this data set, only 1.3% of samples were prepaid in the sampled month. Even looking at distributions conditioned on `prepaid` will be tricky. Let's see if density graphs look ok.

```{r}
sample_data %>% ggplot(aes(x=burnout, col=as.factor(prepaid))) + geom_density()
```
Let's look at continuous variables this way, then.

```{r, fig.height=8, fig.width=10}
#Start with doubles.
cols_dbl <- sapply(spec(sample_data)$cols,function(X){TRUE %in% grepl("double", class(X))})
cols_dbl["prepaid"] = T
sample_data[,cols_dbl] %>% gather(beta,value,-prepaid) %>% ggplot(aes(y=value,x=factor(prepaid))) + geom_boxplot() +
  facet_wrap(~beta, ncol=5, scales="free") 
```

Not seeing a lot of difference in these factors. Also some things are doubles when they clearly shouldn't be, e.g. `property_type` and `original_loan_term`. Let's look at the next set. 
```{r, fig.height=8, fig.width=10}
cols_int <- sapply(spec(sample_data)$cols,function(X){TRUE %in% grepl("integer", class(X))})
cols_int["prepaid"] = T
sample_data[,cols_int] %>% gather(beta,value,-prepaid) %>% ggplot(aes(y=value,x=factor(prepaid))) + geom_boxplot() +
  facet_wrap(~beta, ncol=5, scales="free") 
```

*Nobody* with annual mortgage insurance paid off the loan? That's kind of weird.
```{r}
sample_data %>% filter(prepaid==1) %>% select(annual_mip) %>% max()
```

Could that be a data problem?

`months_prepaid` should definitely be in there. So should `seasonality`. Steve wanted `original_principal_balance`, and there might be something there.

Now, what about categorical variables? I see maturity date was reconstituted poorly. That's in the underlying data, too. Bugger.
```{r, fig.height=8, fig.width=10}
cols_others <- sapply(spec(sample_data)$cols,function(X){!(TRUE %in% grepl("(integer|double|date)", class(X)))})
cols_others['prepaid'] <- T
cols_others['pool_cusip'] <- F

sample_data[,cols_others] %>% gather(beta,value,-prepaid) %>% na.omit() %>% group_by(beta, prepaid) %>%
  do(as_tibble(table(.$value))) %>% group_by(beta, prepaid) %>% mutate(n=n/sum(n)) %>%
  ggplot(aes(y=Var1, x=factor(prepaid), fill=n)) + geom_tile() + facet_wrap(~beta, ncol=4, scales="free") +
  scale_fill_continuous_tableau()
```

Not much difference here. There's an argument for `refinance_type` (which is probably a rare field) , `first_time`, and `agency`. Agency could be a group.

So, let's do a model with:

* burnout
* cato
* sato
* credit_score
* hpa
* incentive
* lockin
* months prepaid
* first_time
* refinance_type (if any)
* seasonality (winter/summer)
* total debt/expense ratio

We can do partial pooling by issuer, but that's a lot of pools. Should also check vintage.